---
title: "Homework 0: Installing R"
author: ""
date: "Last edited `r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
urlcolor: blue
---

-   Due Monday, September 9th, 5:00 pm EDT
-   Submit `hw0.Rmd` and `hw0.pdf` to canvas.
-   Please post questions and discussion you have related to this homework on the Homework 0 discussion thread on Ed.

## Overview

This assignment will show you how to write a report using `R` and Rmarkdown.

HW 0 might be challenging, especially if this is your first time using `R`. First, you'll need to install several pieces of software. You will also need to read and modify some `R` code.

If you are already comfortable using `R` and Rmarkdown then this assignment will be simple. Download `hw0.Rmd`, modify one line of code, then Knit to `.pdf`. This "write then render" workflow is central to making homework submissions using Rmarkdown.

In this course, you aren't required use `R` exclusively. Nevertheless, we will sometimes use `R`, and in the `R` help sessions, we will discuss how ideas from lecture can be put into practice using `R`. Having some understanding of `R` will help you to follow these discussions.

# Complete the class survey on Canvas

Please complete the course survey that is available on Canvas [(link)](https://princeton.instructure.com/courses/16198/quizzes/18052). The purpose of this survey is to assess your level of expertise in programming which will be useful for future assignments and help sessions. The survey will be graded only for completion.

# Installing R and Rstudio

Here is a walk-through of installing R and Rstudio [(link)](https://rstudio-education.github.io/hopr/starting.html). Rstudio is currently the most popular integrated development environment (IDE) for working with R. We recommend that you use it in order to get started quickly. You'll need to install R before installing Rstudio.

#### If you get stuck installing R

Getting stuck on this can be frustrating, but the TAs will help you troubleshoot. Please post the problem you are running into on the Ed discussion thread.

Alternatively, you can complete this assignment without installing R on your PC. Rstudio cloud can run R through a web browser [(link)](https://rstudio.cloud/). You can set up an account for free.

# Navigating the Rstudio IDE

By default, the Rstudio IDE is split into several panes. You'll likely spend most of your time working with the *source editor*, which is the upper left-hand pane. From here you can write and evaluate R. To evaluate a line in the source editor type CTRL-Enter. This will send the line to the *console* and evaluate it. The *console* is below the *source editor*.

Depending on what you're doing in the *source editor* you may want to refer to some of the other panes.

-   **getting help:** you can show documentation in the lower right corner. For example, evaluating `?mean` will show the documentation for the `mean` function.
-   **refining a plot:** You can display a plot in the lower right-hand pane by clicking on the plots tab. The arrows at the top of this pane allows you to scroll between recent plots. For many R programmers, iterating back and forth between writing R and plotting is an important part of the workflow.
-   **looking up variables:** If you want to check what a variable is named or get info on how it is structured, you can use the *environment* tab. `Environments` are used by R to associate names to values, hence the two columns of the *environment* tab.

# Reproducible reports using Rmarkdown

In this class you'll be writing reports in Rmarkdown. Some helpful instructions include:

-   a tutorial by RStudio [(link)](https://rmarkdown.rstudio.com/lesson-1.html).
-   a more concise guide [(link)](https://rmarkdown.rstudio.com/authoring_basics.html).

You can also use the `.Rmd` that generated this homework as a template.

Rmarkdown includes a light-weight set of markup elements that allow you to do some basic formatting of your document. For example, you can easily specify

-   **bold text**,
-   and lists.

The key feature of Rmarkdown is the ability to incorporate code and computation into your writing.

## Installing Rmarkdown

To compile an Rmarkdown document you need to install the `rmarkdown` package in `R`. This package has its own set of dependencies including `latex` and `pandoc`. If you are using Rstudio, it will automatically install `pandoc`. Otherwise, you'll need to install these dependencies yourself. *R markdown: the definitive guide* includes a chapter with instructions for installing `rmarkdown` [(link)](https://bookdown.org/yihui/rmarkdown/installation.html).

## Rendering an Rmarkdown document

Once you have the `rmarkdown` package installed you have a few options for rendering an Rmarkdown document.

-   If you use the Rstudio IDE then you can simply click a button to render your document. This button looks like a ball of yarn and is above the source editor, make sure you select **Knit to PDF** from the available options. The shortcut to render is Ctrl+Shift+K.
-   Alternatively, you can do this from the command line. First move to the directory that contains your Rmarkdown source code. Next render the document using something like

```         
Rscript -e "rmarkdown::render('./hw0.Rmd')"
```

This assignment was written using a `.Rmd` file. The later parts of the assignment will ask you to modify it. Before making changes you should download the `.Rmd` from canvas and render it to produce a `.pdf` file.

# (Homework begins here) Edgar Anderson's Iris Data

We'll begin by working with the famous (Fisher's or Anderson's) iris data.

You can access a description of this data and its format using the help function in R: **?**.

```{r, eval=FALSE}
?iris
```

> Description

This famous (Fisher's or Anderson's) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica.

> Format

iris is a data frame with 150 cases (rows) and 5 variables (columns) named Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, and Species.

iris3 gives the same data arranged as a 3-dimensional array of size 50 by 4 by 3, as represented by S-PLUS. The first dimension gives the case number within the species subsample, the second the measurements with names Sepal L., Sepal W., Petal L., and Petal W., and the third the species.

### Loading and examining data

In this assignment we will give an example of scraping, formatting, and previewing external data from an URL. Just like with **?iris**, you can load a description of any of these functions using the help function. (**?read.csv**, **?head**, **?summary**)

```{r}
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/iris/iris.data"
iris_data <- read.csv(url, header = FALSE, col.names = c("Sepal.Length", 
                                                         "Sepal.Width", "Petal.Length", 
                                                         "Petal.Width", "Species"))
# Preview the first 5 rows
head(iris_data, 5)

# Summarize each column of the dataframe
summary(iris_data)

iris3 <- iris3
```

### Plotting the data

Finally, we can visualize our data using the plot() function (**?plot.default** for info!). We'll specify data for each axis using **\$**, which allows us to specify and select a column from our data object. Here we plot the distribution of Sepal Length (**iris_data\$Sepal.Length**) and Sepal Width (**iris_data\$Sepal.Width**) between the species of the iris flower (setosa, versicolor, and virginica).

```{r}
# Create a scatter plot with a legend, axis labels, and title
plot(iris_data$Sepal.Length, iris_data$Sepal.Width, col = as.factor(iris_data$Species),
     main = "Sepal Length vs Sepal Width by Species",
     xlab = "Sepal Length (cm)",
     ylab = "Sepal Width (cm)")

# Add a legend to the plot
legend("topleft", legend = levels(as.factor(iris_data$Species)), 
       col = 1:3, pch = 1, cex = 0.8)
```

### HOMEWORK: Modify here.

Now, modify the code below to show the distribution of Petal Lengths and Widths across the 3 iris species. The plot below is again just currently plotting the same values on the y-axis as on the x-axis.

```{r}
# TODO: Modify this statement
plot(iris_data$Petal.Length, iris_data$Petal.Width, col = as.factor(iris_data$Species),
     main = "Petal Length vs Petal Width by Species",
     xlab = "Petal Length (cm)",
     ylab = "Petal Width (cm)")
legend("topleft", legend = levels(as.factor(iris_data$Species)), 
       col = 1:3, pch = 1, cex = 0.8)
```

#### Submitting the assignment

You can complete this assignment by modifying only the lines marked `TODO` above. Once you modify the plot above, render your `.Rmd` file to produce a `.pdf`. There are instructions for doing this in the *Rendering an Rmarkdown document* section above. Submit both your `.Rmd` and `.pdf` file to canvas.

## References

Anderson, Edgar (1935). The irises of the Gaspe Peninsula, Bulletin of the American Iris Society, 59, 2–5.
